### 2016-07-03 case-studies.tex
Bib fixed. Case studies edited.

\paragraph{A video view counter} The view count of a video on a
YouTube-like video-sharing website must never appear to be decreasing,
although it may not necessarily reflect the actual number of views.
We call this the monotonicity invariant. To understand whether a
straightforward implementation of the view counter preserves the
invariant, we considered a \txnimp program that is representative of
the counter application. In particular, we considered four concurrent
transactions that operate on a shared counter: a reader transaction
that reads the counter twice obtaining the values $k_1$ and $k_2$,
respectively, and three writer transactions, each incrementing the
counter once. If the reader witnesses a monotonically increasing view
count, then $k_2\ge k_1$. We then instantiated our reasoning framework
for an {\sc sc} store with {\sc rc} as the default isolation level
(Postgres is one such store), with the aim of verifying the program
under this instantiation.  Our attempts were however unsuccessful
owing to our inability to prove that a writer transaction always
commits a value no smaller than the current value of the counter. This
points to the existence of anamolous executions where writers may
commit a value smaller than the current value of the counter. To
understand the prevalence of such anomalies, we implemented a view
counter application in Ruby-on-Rails using standard templates and
techniques\footnote{Source code of the applications, experiments, and
the tools we developed to analyze Ruby can be found at the following
anonymized url: \url{...}}. The application stores the video count in
a Postgres database, and supports an increment operation and a read
operation on the count. In our experiments with 16 (resp. 32)
concurrent writers, readers observed 7 (17) violations of
monotonicity, in 10 rounds of operation, thus confirming that
anomalies are indeed possible.

To determine the weakest ANSI SQL isolation levels needed to restore
the monotonicity invariant, we instantiated our reasoning framework
for an {\sc sc} store with different isolation levels for readers and
writers.  With half-a-day of effort, we were able to prove that
\iso{Read Uncommmitted}, the weakest isolation level, is sufficient
for readers, whereas \iso{Serializable} isolation is needed for
writers. Our experiments with uncommitted reads and serializable
writes resulted in no observed violations of monotonicity, confirming
our judgment. While this exercise let us determine the appropriate
isolation levels for an {\sc sc} store with ANSI SQL isolation levels,
we were however surprised to learn from Postgres's
documentation~\cite{postgres} that its isolation levels are stronger
than ANSI SQL's in non-trivial ways. We subsequently formalized the
guarantees and repeated the proof effort to discover that Postgres's
\iso{Repeatable Read} is in fact sufficient for writers - an
observation we confirmed through experiments. However, the same choice
of isolation levels led to monotonicity violations on MySQL, because
MySQL's {\sc rr} is weaker than Postgres's.

To understand the effect of changing the store consistency, we
reimplemented the counter application for Cassandra, an {\sc ec} store
(We used the \emph{shim layer} of~\cite{pldi15} to equip Cassandra
with {\sc cc} and {\sc sc} consistency levels, and {\sc rc}, {\sc
mav}, and {\sc rr} isolation levels). We then repeated our experiments
with uncommitted reads and serializable writes - a combination which
was proved to preserve monotonicity on an {\sc sc} store, but
nonetheless observed monotonicity violations. The reason for such
violations became evident as we considered an instantiation our proof
framework for an {\sc ec} store. Under this instantiation, it is
impossible to derive the \C{MonotonicVis} (\S\ref{sec:ansi-isolation})
property needed to show that successive reads in a session witness
monotonically increasing state. We were however able to derive the
property assuming either causal consistency or {\sc mav} isolation
(both supported by the shim layer). We experimented with both these
options and were able to recover the invariant.

@@

Next, we focus our attention on a Twitter-like microblogging site that
is one the sample applications shipped with a popular textbook on
Ruby-on-Rails~\cite{railsbook}. The application is built to
demonstrate various capabilities of Rails' ORM framework, and, as
such, is a good representative of Ruby-on-Rails applications.  For
instance, instead of relying on database capabilities, it uses Rails's
\emph{has\_many} association to describe the relationship between a
user entity and a micropost entity. Further, it qualifies the
association with \emph{dependent: destroy}, thus directing Rails to
destroy all microposts if the corresponding author is deleted. Thus,
the application entrusts Rails with the responsibility to enforce
referential integrity between microposts and users. Rails in turn
relies on database-backed transactions to enforce such constraints,
thus making application semantics susceptible to the changes in
isolation level.  For instance, deletion of a user is done within a
transaction ($T_1$) that also deletes the microposts authored by
(hence, \emph{dependent} on) the user. While this is effective if all
transactions are {\sc ser}, under the default {\sc rc} isolation, it
lets concurrent transactions witness referential integrity violations.
For instance, consider a transaction ($T_2$) that populates a timeline
by first reading microposts, and then reading the author information
associated with each micropost. It is possible for an author to be
deleted (due to the interference from $T_1$) between the two reads,
leading $T_2$ to witness a referential integrity violation, although
none really exists in the database.  Indeed, we have verified this
hypothesis by instantiating our reasoning framework for {\sc sc}
machine and {\sc rc} isolation, and extracting the aforementioned
anomalous execution from our failure to prove referential
integrity\footnote{We modeled referential integrity using two
variables, $X$ and $Y$: $Y$ is 1 only if $X$ is 1.  Otherwise, both
are 0.}. To measure the likelihood of such anomalies in practice, we
first populated a Postgres database with 1000 user accounts, each user
with 50 microposts and 20 followers, choosen uniformly at random. We
then constructed experiments with 64 concurrent clients performing
deletions and timeline reads of randomly choosen users in 1:7 ratio,
and witnessed a maximum of 250 violations of referential integrity.

Clearly, {\sc rc} is insufficient. In order to determine the weakest
isolation levels needed to enforce referential integrity, we
instantiated our framework for an {\sc sc} store with various Postgres
isolation levels for user deletions and timeline reads, and obtained a
proof that {\sc rc} is sufficient for deletions, whereas reads require
{\sc rr}. Repeating the experiment with these isolation settings
resulted in no violations, as expected.

>>>>

\paragraph{A video view counter} Our first application is a
Ruby-on-Rails implementation of a counter to count the number of views
a video has garnered on a video-sharing website like
YouTube\footnote{Source code of the applications, experiments, and the
tools we developed to analyze Ruby can be found at the following
anonymized url: \url{...}}. The application stores the video count in
a Postgres database, and supports a read operation (a database read
transaction), and an increment operation (a read-increment-write
sequence wrapped in a transaction). The application is expected to
preserve the \emph{monotonicity invariant}, namely that the view count
should never appear to be decreasing. However, the default {\sc rc}
isolation of Postgres may lead to anomalous executions as confirmed by
our experiments; with 16 (resp.  32) concurrent writers performing
single increments, we observed a total of 7 (17) violations of
monotonicity. We subsequently instantiated our reasoning framework for
an {\sc sc} store with Postgres isolation levels (which are stronger
than ANSI SQL's and MySQL's), and spent 5 man hours determining
appropriate isolation levels needed to enforce the
invariant\footnote{The representative \txnimp programs we used to
formally reason about scenarios described in this section can be found
in the supplementary.} In particular, we were able to formally prove
that to preserve the invariant, writers need to execute at
\iso{Repeatable Read} isolation, while readers can continue at {\sc
rc}. Repeating the experiments with this configuration led to no
violations, as expected. While executing all transactions at {\sc ser}
also prevents violations, it led to an additional 47\% increase in
latency on an average (16 concurrent writers; 5 rounds). We repeated
the exercise for MySQL, and found that its {\sc rr} is not sufficient
to enforce monotonicity; {\sc ser} is needed.

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency, we
reimplemented the counter as a \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels, and ANSI SQL isolation levels. Keeping the
consistency level at {\sc ec}, we then repeated our experiments with
committed reads and serializable increments - a combination which was
proved to preserve monotonicity on an {\sc sc} store, but nonetheless
observed monotonicity violations. On the other hand, instantiating the
store consistency with {\sc cc} allowed us to obtain a proof, and
recover the monotonicity invariant. We in fact obtained a proof after
relaxing the serializability requirement for writes, by changing the
definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$.
We implemented this version of the counter RDT, and obtained a 52\%
improvement in latency compared to the previous
implementation\footnote{The tradeoff is that the view count is now no
longer accurate, although it grows monotonically.}.

@@

Next, we focus our attention on a Twitter-like microblogging site that
is one the sample applications shipped with a popular textbook on
Ruby-on-Rails~\cite{railsbook}. One of the invariants that the
application chooses to enforce is the referential integrity between
microposts and authors. The application enforces this invariant
entirely in the user space by relying on database-backed transactions
(this behaviour is typical of Rails applications~\cite{bailisferal}).
For instance, deletion of a user is carried out in a transaction that
also includes the deletion of microposts authored by the user.
However, concurrent observers can nonetheless witness violations of
referential integrity. To measure the likelihood of such anomalies in
practice, we first populated a Postgres database with 1000 user
accounts, each user with 50 microposts and 20 followers, choosen
uniformly at random. We then constructed experiments with 64
concurrent clients performing deletions and timeline reads of randomly
choosen users in 1:7 ratio, and witnessed a maximum of 250 violations
of referential integrity. We then instantiated our framework for an
{\sc sc} store with various ANSI SQL isolation levels for user
deletions and timeline reads, and obtained a proof that {\sc rc} is
sufficient for deletions, whereas reads require {\sc rr}. The proof is
applicable for MySQL, which implements ANSI isolation.  Since
Postgres's isolation levels are stronger than corresponding ANSI
levels, using \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}), we
also obtain a correctness proof for Postgres.

As evident from our case studies, the there already exists a
significant diversity among the weak consistency and isolation
variants implemented on various real-world data stores. More such
variants have been developed in the recent years to meet the need for
high availability on geo-replicated stores. For example,
\emph{Parallel Snapshot Isolation}~\cite{psi} was proposed in 2011,
and \emph{Non-Monotonic Snapshot Isolation}~\cite{nmsi} in 2013, both
targeted for replicated stores. As more commercial weakly consistent
stores adopt various kinds of transactional semantics, like the
recently introduced \emph{lightweight transactions} of Cassandra, it
is reasonable to expect new isolation definitions to continue to be
proposed. It may be possible to carefully engineer a reasoning
framework for each combination of consistency and isolation, but such
a strategy would lead to multiple semantic definitions and proof
systems with no obvious way to compare and relate them. Having a
parameterized semantics and a proof framework built on the such
semantics allows us to support these variants as distinct, yet
comparable instantiations, as we demonstrate in our case studies.
%  GK: what does the following sentence mean?
%  As a consequence, we avoid the need for semantics re-engineering
%  every time an application is retargeted to a new platform.

### 2016-07-03 introduction.tex
Bib fixed. Case studies edited.

A major problem with weak isolation is that its semantics in the
context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard defines
four levels of isolation (including serializability) in terms of
various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
required to prevent. While this is an improvement, it requires
programmers to be prescient about the possible ways various
undesirable phenomena might manifest in their applications, and
determine, in each case, if the phenomenon can be allowed without
violating application invariants. This is understandably hard,
especially in the absence of any formal underpinning to define weak
isolation semantics.  Adya~\cite{adyaphd} presents the first formal
definitions of some of well-known isolation levels in the context of a
sequentially consistent (SC) database.  However, there has been little
progress relating Adya's system model to a formal operational
semantics or a proof system that can facilitate rigorous correctness
arguments.  Consequently, reasoning about weak isolation remains an
error prone endeavor, with major database vendors continuing to
document their isolation levels primarily in terms of the undesirable
phenomena~\cite{postgresiso, mysqliso, oracleiso} a particular
isolation level may induce, placing the burden squarely on the
programmer to determine application correctness.

>>>>

A major problem with weak isolation is that its semantics in the
context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard defines
four levels of isolation (including serializability) in terms of
various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
required to prevent. While this is an improvement, it requires
programmers to be prescient about the possible ways various
undesirable phenomena might manifest in their applications, and
determine, in each case, if the phenomenon can be allowed without
violating application invariants. This is understandably hard,
especially in the absence of any formal underpinning to define weak
isolation semantics.  Adya~\cite{adyaphd} presents the first formal
definitions of some of well-known isolation levels in the context of a
sequentially consistent (SC) database.  However, there has been little
progress relating Adya's system model to a formal operational
semantics or a proof system that can facilitate rigorous correctness
arguments.  Consequently, reasoning about weak isolation remains an
error prone endeavor, with major database vendors~\cite{postgresiso,
mysqliso, oracleiso} continuing to document their isolation levels
primarily in terms of the undesirable phenomena a particular isolation
level may induce, placing the burden squarely on the programmer to
determine application correctness.

### 2016-07-03 store-consistency.tex
Bib fixed. Case studies edited.

\paragraph{Discussion.} As evident from the example, instantiating the
proof system for a specfic combination of consistency and isolation
settings, and then deriving facts specific to that setting is one of
the characteristic our reasoning approach. While intuition suggests
that some instantiations, such as {\sc sc} with ANSI SQL isolation,
are general enough to cover many practical cases, it not the case in
reality owing to the subtle differences in the semantics of isolation
levels among databases. For instance, Postgres's \iso{Repeatable Read}
isolation is stronger than ANSI SQL's or MySQL's in non-trivial ways.
Furthermore, the need for high availability on geo-replicated stores
has led to the development of of new isolation levels which are yet to
be standardized. For example, \emph{Parallel Snapshot
Isolation}~\cite{psi} was proposed in 2011, and \emph{Non-Monotonic
Snapshot Isolation}~\cite{nmsi} in 2013, both targeted for replicated
stores. As more commercial weakly consistent stores adopt variants of
transactions (like the \emph{lightweight transactions} of Cassandra),
it is reasonable to expect novel forms of isolation to be proposed.
While it is possible to carefully engineer a reasoning framework for
each combination of consistency and isolation, it leads to multiple
proof systems with no obvious way to compare proofs.  Having a
parameterized semantics and proof framework built on this semantics
allows us to support these variants as distinct, yet comparable
instantiations, thus avoiding the need for semantics re-engineering
every time an application is retargeted to a new platform

>>>>

% \paragraph{Discussion.} As evident from the example, instantiating the
% proof system for a specfic combination of consistency and isolation
% settings, and then deriving facts specific to that setting
% characterizes our reasoning approach. While intuition suggests that
% some instantiations, such as {\sc sc} with ANSI SQL isolation, are
% general enough to cover many practical cases, it not the case in
% reality given the subtle differences in the semantics of isolation
% levels among different database implementations. For instance,
% Postgres' \iso{Repeatable Read} isolation is stronger than ANSI SQL's
% or MySQL's in non-trivial ways.  Furthermore, the need for high
% availability on geo-replicated stores has led to the development of
% new isolation levels which are yet to be standardized. For example,
% \emph{Parallel Snapshot Isolation}~\cite{psi} was proposed in 2011,
% and \emph{Non-Monotonic Snapshot Isolation}~\cite{nmsi} in 2013, both
% targeted for replicated stores. As more commercial weakly consistent
% stores adopt variants of transactions, like the recently introduced
% \emph{lightweight transactions} of Cassandra, it is reasonable to
% expect new isolation definitions to continue to be proposed.  While it
% is possible to carefully engineer a reasoning framework for each
% combination of consistency and isolation, such a strategy would lead
% to multiple semantic definitions and proof systems with no obvious way
% to compare and relate them.  Having a parameterized semantics and a
% proof framework built on the such semantics allows us to support these
% variants as distinct, yet comparable instantiations, thus avoiding the
% need for semantics re-engineering every time an
% application is retargeted to a new platform.
